---
title: "Mitochondrial gene expression in lung cancer: Significantly overexpressed genes in LUSC"
output: pdf_document
---

Report created: `r Sys.time()`

This document summarizes analysis of TCGA data from lung cancer projects to assess expression patterns and survival among lung cancer patients.

LUAD: https://www.nature.com/articles/nature13385

Please compare each significantly overexpressed gene with stage and grade and smoking, EGFR/KRAS/LKB1 expression

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse)
```

```{r gene_data, include=FALSE}
# read in data
fpkm <- read_csv(stringr::str_c("TCGA-LUAD.csv"))
```

### LUSC: Does overexpression vary with stage?

LUSC: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3466113/

```{r}
# read in gene significance
LUSC_pval_ttest <- read_csv("mt-LUAD-ttest-sig.csv")
# show list of only significantly overexpressed genes
LUSC_pval_ttest <- LUSC_pval_ttest %>%
  filter(ttest_sig == TRUE, overexpressed == TRUE)
```

```{r}
# tumor stage data
#table(fpkm$tumor_stage) # includes "not reported"
table(fpkm$ajcc_pathologic_stage)
# group into Stage I, II, III/IV
fpkm <- fpkm %>%
  # remove subcategories
  collapse_stages(ajcc_pathologic_stage) %>%
  # combine III and IV (to Stage III)
  mutate(collapsed_stages = str_replace_all(collapsed_stages, "Stage IV", "Stage III"))
table(fpkm$collapsed_stages)
# reformat data and plot
fpkm %>%
  dplyr::select(LUSC_pval_ttest$mt_gene, collapsed_stages) %>%
  dplyr::filter(!is.na(collapsed_stages)) %>%
  pivot_longer(cols = LUSC_pval_ttest$mt_gene,
               names_to = "gene", values_to = "expression") %>%
  ggplot() +
    geom_boxplot(aes(collapsed_stages, expression)) +
    facet_wrap("gene") +
    labs(y = "log2 expression",
         x = "stage") +
    scale_x_discrete(
      labels = c("Stage I"="I", "Stage II"="II", "Stage III"="III/IV")) +
    theme_bw()
```

### LUSC: Does overexpression vary with smoking?

fpkm$paper_Smoking.Status
