---
title: "Mitochondrial gene expression in lung cancer: `r params$gene` in `r params$cancer`"
output: pdf_document
params: 
  gene: "SLC25A4"
  cancer: "LUSC"
---

Report created: `r Sys.time()`

This analysis uses TCGA data from the `r params$cancer` project to assess `r params$gene` expression patterns and survival among lung cancer patients.

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, survival, survminer)
library(tcgaHelpers)
```

```{r load_data, include=FALSE}
fpkm <- read_csv(stringr::str_c("TCGA-", params$cancer, ".csv"))
```

### Overall expression pattern

```{r tumor_exp, echo=FALSE, warning=FALSE, fig.height=3, fig.width=3}
# assess expression for tumor vs normal
fpkmTum <- fpkm %>%
  select(params$gene, shortLetterCode, vital_status, days_to_death) %>%
  filter(shortLetterCode == "NT" | shortLetterCode == "TP") %>%
  # code vital status for survival analysis 
  mutate(vital = as.numeric(as_factor(vital_status)) - 1)
# rename target gene column
names(fpkmTum)[1] <- "test_gene"
# adding column for gene expression categories
fpkmTum <- fpkmTum %>%
  mutate(gene_hl = test_gene)
# perform t-test
tum.results <- t.test(test_gene ~ shortLetterCode, data = fpkmTum) 
# determine overexpressed status
overexpress <- tum.results[["estimate"]][["mean in group NT"]] < tum.results[["estimate"]][["mean in group TP"]]
# boxplot to compare distributions
ggplot(fpkmTum, aes(shortLetterCode, test_gene)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.3) +
  labs(y = "log2 expression",
       x = "sample") +
  scale_x_discrete(labels = c("NT"="normal", "TP"="tumor")) +
  theme_bw() 
```

Does `r params$gene` expression differ between normal
(n = `r table(fpkmTum$shortLetterCode)[1]`) and 
primary tumor (n = `r table(fpkmTum$shortLetterCode)[2]`) (t-test),
and is it overexpressed?

p = `r tum.results$p.value`

overexpressed = `r overexpress`

```{r hi_lo_median, eval=FALSE, include=FALSE}
## assessing low and high expression based on median
fpkmTum <- fpkmTum %>%
  mutate(gene_hl = test_gene > median(test_gene))
fpkmTum$gene_hl[fpkmTum$gene_hl == TRUE] <- "high"
fpkmTum$gene_hl[fpkmTum$gene_hl == FALSE] <- "low"
```

The cutpoint (threshold between high and low gene expression)
is empirically determined by the data:

```{r hi_lo, echo=FALSE, message=FALSE}
# determine cutpoint
cut <- surv_cutpoint(fpkmTum, time = "days_to_death", event = "vital",
   variables = "gene_hl")
#summary(cut)
# show plot
cut_plot <- plot(cut)
cut_plot$gene_hl
# categorize based on cutpoint
fpkmSurv <- surv_categorize(cut)
```

```{r histogram, echo=FALSE, message=FALSE}
ggplot(fpkmTum) + 
  geom_histogram(aes(test_gene), bins = 50) +
  geom_vline(xintercept = cut$cutpoint$cutpoint) +
  labs(title = "Distribution of expression in tumor samples") +
  theme_bw()
```

## Survival analysis

```{r survival}
# fit model
death_fit <- survfit(Surv(days_to_death, vital) ~ gene_hl, 
                               data = fpkmSurv)
# complete plot
survPlot <- ggsurvplot(death_fit, 
           data = fpkmSurv, 
           risk.table = TRUE,
           pval = TRUE,
           pval.method = TRUE,
           pval.coord = c(2000, 0.8),
           pval.method.coord = c(2000, 0.9),
           conf.int = TRUE,
           xlim = c(0, 5000),
           break.time.by = 500,
           xlab = "days to death",
           ggtheme = theme_bw(),
           risk.table.y.text.col = TRUE,
           risk.table.y.text = FALSE,
           legend.labs = c("high", "low"))
survPlot
```

```{r survival_fig}
# plot for figure
survPlot <- ggsurvplot(death_fit, 
           data = fpkmSurv, 
           conf.int = TRUE,
           xlab = "",
           ylab = "",
           ggtheme = theme_bw(),
           legend = "none",
           palette = c("#7fc97f", "#beaed4")) # green=hi, purple=lo
```

```{r surv_diff}
# calculate difference
death_diff <- survdiff(Surv(days_to_death, vital) ~ gene_hl, 
                               data = fpkmSurv)
# calculate p-value
p_diff <- 1 - pchisq(death_diff$chisq, length(death_diff$n) - 1)
```

```{r save_results}
# save results to summary file
write_csv(tibble(mt_gene = params$gene,
                 pvalue_ttest = tum.results$p.value,
                 overexpressed = overexpress, 
                 pvalue_surv = p_diff),
          file = stringr::str_c("mt-", params$cancer, "-pvalue.csv"),
          append = TRUE)
```

