---
title: "Mitochondrial gene expression in lung cancer: `r params$cancer` summary"
output: pdf_document
params:
  cancer: "LUSC"
---

Report created: `r Sys.time()`

This document summarizes analysis of TCGA data from lung cancer projects to assess expression patterns and survival among lung cancer patients.

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, AnnotationDbi, org.Hs.eg.db)
```

```{r data, include=FALSE}
# read in data
pval <- read_csv(stringr::str_c("mt-", params$cancer,"-pvalue.csv"))
# remove (empty) first row
pval <- pval[-1, ]
```

### `r params$cancer`: Does expression differ between tumor and normal?

```{r ttest, message=FALSE, results='asis'}
# correct for multiple tests
pval <- pval %>%
  mutate(pval_ttest_adj = p.adjust(pvalue_ttest, method = "bonferroni")) %>%
  mutate(pval_ttest_adj = signif(pval_ttest_adj, digits = 6)) %>%
  mutate(ttest_sig = pval_ttest_adj < 0.05)
pval_ttest <- pval %>%
  dplyr::select(mt_gene, pvalue_ttest, pval_ttest_adj, ttest_sig, overexpressed) %>%
  arrange(mt_gene)
knitr::kable(pval_ttest)
```

```{r save_ttest}
# save final table to file
write_csv(pval_ttest, file = stringr::str_c("mt-", params$cancer, "-ttest-sig.csv"))
```

### `r params$cancer`: Does survival differ between high and low expression?

**Important note**: The p-values that appear on the survival plots for individual genes have not been corrected for multiple comparison.

```{r surv, message=FALSE, results='asis'}
# correct for multiple tests
pval <- pval %>%
  mutate(pval_surv_adj = p.adjust(pvalue_surv, method = "bonferroni")) %>%
  mutate(pval_surv_adj = signif(pval_surv_adj, digits = 6)) %>%
  mutate(surv_sig = pval_surv_adj < 0.05)
# create table for printing
pval_surv <- pval %>%
  # remove samples that were not significantly different in previous analysis
  filter(ttest_sig == TRUE) %>%
  dplyr::select(mt_gene, pvalue_surv, pval_surv_adj, surv_sig) %>%
  arrange(mt_gene)
knitr::kable(pval_surv)
```

Only the `r nrow(pval_surv)` genes which differed significantly between tumor and normal samples were included in this analysis.

```{r save_surv}
# save final table to file
write_csv(pval_surv, file = stringr::str_c("mt-", params$cancer, "-surv-sig.csv"))
```
